\Large \bf Vigor: Toward Carrier-Grade Software Network Functions

### 2016-11-17 doc/paper/paper.tex
Refactor the solution overview section and divide it into two parts: developer experience and "under the hood"

@@

In order to absorb bursts, the code employs a ring buffer. The interface to the
network is: \code{bool receive_packet(struct packet* dst)}, which non-blockingly
reads an inbound packet and stores it at \code{dst}, returning success or
failure; \code{bool can_send_packet()}, which checks if a new packet can be
sent; and \code{void send_packet(struct packet* src)}, which sends the packet
stored at \code{src}. The application follows common practice: it consists of a
main event loop, it has no other complex loops, and it uses a ring, a popular
data structure in network programming.

One could use an SEE like KLEE~\cite{cadar2008klee} to symbolically execute the
code above exhaustively --- once all paths are found to be crash-free, the
crash-freedom proof is complete. To perform SE, it is necessary to provide
symbolic models for the three network calls. Something simple suffices, such as
\code{receive_packet} generating a new symbolic packet in \code{dst};
\code{can_send_packet} forking program state into two states for busy and idle
interface correspondingly; and \code{send_packet} doing nothing.

Alas, symbolic execution would never terminate. The reason is ``path
explosion'': there is a large (often infinite) number of paths to explore, so
the SEE cannot finish. Path explosion results from cyclomatic
complexity~\cite{mccabe:cyclomatic}: every branch condition doubles the number
of paths, and every loop multiplies it by a factor; in our case, the infinite
loop gives rise to an infinite number of paths.

{\bf Step 1: Add loop invariants.} Infinite (or very long) loops can be
simplified using loop invariants~\cite[\S~2.1]{cormen2009introduction}. These
are annotations that inform the SEE about invariant properties of the loop, such
as ``no element of \code{packets} between \code{begin} and \code{begin +
  queue_len} has \code{port == 9}'', and are a good match for NF-style loops. We
added to KLEE support for havoc-ing~\cite{barnett2005boogie} automatically
(replace all modified variables with unconstrained symbols), to take advantage
of such invariants and essentially reduce the infinite loop to a single abstract
iteration that allows reasoning in one pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. It takes so long due to another classic
SE challenge, ``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset}: when
indexing an array with a symbolic variable (or when dereferencing a symbolic
pointer), the constraints on program state become extremely complex, because the
symbolic pointer or index could be referencing many different possible
locations. The complexity of the constraints chokes the SMT solver, and the SEE
grinds to a halt. In general, SE doesn't work for code with symbolic pointers or
offsets. Alas, stateful C code almost always has some of these.

{\bf Step 2: Use abstract data structures.} In Vigor, we eliminate the problem
of symbolic indexing by encapsulating program state into well-defined data
structures and putting them into a ``library of components''. This works well
for NF development, because the NF code uses a relatively small set of standard
structures, such as ring buffers, hash tables, least-prefix-match tables, etc.
We provide to the SEE symbolic models of these abstracted components that
``hide'' symbolic indexes and pointers from the SEE.

For our example, the library provides a packet ring with a simple interface:
\code{bool ring_empty(ring*)} checks whether there is a packet in the ring,
\code{bool ring_full(ring*)} checks whether the ring has run out of space, \code{void
  ring_push_back(ring*, packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(ring*, packet* dst)} removes a packet from the ring
and places it in \code{dst}. We then write a symbolic model for the \code{ring}
and provide it to the SEE instead of the actual implementation. The NF now
becomes:

@@

For the symbolic execution and future validation, we need also to customize the
general-purpose packet ring library with the following definitions:

@@

Currently, we have to duplicate the packet condition in two languages (formal
logic, and C) due to the limitations of our prototype. This is the
application-specific part of the loop invariant given before, that we plug-in in
order for Vigor to be able to verify the app.

Now exhaustive SE of the code takes less than 100 millisec with our modified
KLEE, regardless of the size of \code{CAP}.  The fact that the code does not
crash on any path amounts to a proof of crash freedom. Or does it? The validity
of the proof depends on (1)  our library implementation of \code{ring} not
crashing; and (2) the symbolic model of \code{ring} accurately capturing how the
\code{ring} implementation works ({\em Step 3}).

Suppose we had an overapproximate model:
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  // No constraints on the packet port here.
}
\end{lstlisting}
  \caption{An extract from an overapproximate ring model.}
  \label{lst:overapproximate}
\end{figure}
Then the \code{VIGOR_CHECK} from the previous listing would fail.

The Vigor library comes with a formal specification and a formal proof that the
implementation satisfies that specification. In other words, there is a formal
spec for \code{ring} and a proof that the \code{ring} implementation satisfies
the spec. This is how we enable SE to verify the high level code, while using
separation logic to verify the library of components---two different
verifications techniques for two different parts of the NF. Doing the library
verification requires deep expertise in program verification, but our effort is
amortized across all NFs that use the library. Vigor makes it possible for NF
developers to verify their code without having to understand verification. Note
that stating invariants for dataplane software turns out to require no
verification expertise.

Here is the example of the proven specification for the function
\code{ring_pop_front}:

@@

\begin{lstlisting}
ringp(r, user_property, tail(lst), cap) &*&
packetp(p, head(lst)) &*&
true == user_property(head(lst))};
\end{lstlisting}
means:
\begin{itemize}
\item \code{r} points to a healthy ring of capacity \code{cap}, that contains
  \code{tail(lst)} packets, that satisfy the \code{user_property}.
\item \code{p} points to a packet \code{head(lst)}.
\item \code{head(lst)} satisfies the \code{user_property}.
\end{itemize}

Note that \code{user_property(head(lst));} in the post condition ensures
\code{p->port != 9} property.

{\bf Step 3: Validate the model used for SE.} The model must match the library
spec. Some symbolic models work well for some code, while others work well for
other code, so the Vigor library comes with {\em multiple} simple,
overapproximating models for each component, and Vigor automatically chooses
combinations of models for symbolically executing the NF code. Once it finds a
combination of models with which the desired property (crash freedom, no memory
leaks, absence of certain security vulnerabilities) can be proven, Vigor
validates that the behaviors exposed by the model during verification are
consistent with the component's formal specification. Details on how it does
this appear in \S\ref{sec:design}.

@@

You can see the example call sequence below. It starts with
\code{loop_invariant_produce} that marks the beginning of a loop iteration. It
continues then with calls to \code{ring_full} \code{ring_empty}, noting (with
\code{assume}) that both return \code{false}. The seqeunce ends with a call to
\code{ring_pop_front}, formally specified above. The last line checks that the
value returned by model (\code{arg2.port != 9}) covers all the cases prescribed
by the specification.

\begin{figure}[h!]
\begin{lstlisting}
struct ring* arg1;
struct packet arg2;

@@

/*@ assert((false == (9 == arg2.port))); @*/
\end{lstlisting}
  \caption{A call sequence, produced by SEE.}
  \label{lst:call-sequence}
\end{figure}

Formal verification engine then checks this sequence using the contracts
like in the Figure~\ref{lst:contract}, and ensures the abscence of misuse. It then checks
that the model output covers all the possible outputs specified in the contract
for the tip function (see~\S\ref{sec:model-validation-criteria}).
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  p->port = 0;
}
\end{lstlisting}
  \caption{An underapproximate model fragment.}
  \label{lst:underapproximate}
\end{figure}

Suppose we had an underapproximate model: Figure~\ref{lst:underapproximate}.
Then the last line in the call sequence check will look like:
\begin{lstlisting}
/*@ assert(0 == arg2.port); @*/
\end{lstlisting}
And the formal verification will fail, because the contract
(Figure~\ref{lst:contract}) specifies a wider range of possible outputs.

{\bf Step 4: Link the NF code with the verified library} of components. The
resulting executable is guaranteed to be free of crashes, subject to the
assumptions listed in~\S\ref{sec:limitations}.

>>>>

\subsection{Developer experience}

@@

In order to absorb bursts, the code employs a ring buffer. The interface to the
network is: \code{bool receive_packet(struct packet* dst)}, which non-blockingly
reads an inbound packet and stores it at \code{dst}, returning success or
failure; \code{bool can_send_packet()}, which checks if a new packet can be
sent; and \code{void send_packet(struct packet* src)}, which sends the packet
stored at \code{src}. The application follows a common practice: it consists of a
main event loop, it has no other complex loops, and it uses a ring, a popular
data structure in network programming.

{\bf Step 1: Label infinite loops.} Our approach requires special handling of
infinite loops in the application code. A developer puts a label \code{LOOP}
into the condition in order to designate the potentially infinite loop. For a
typical NF there should be only a handful or them (1 in our example).

{\bf Step 2: Use library components. } A developer replaces every array with a
component from Vigor library that performs the same function. For our example,
the library provides a packet ring with a simple interface: \code{bool
  ring_empty(ring*)} checks whether there is a packet in the ring, \code{bool
  ring_full(ring*)} checks whether the ring has run out of space, \code{void
  ring_push_back(ring*, packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(ring*, packet* dst)} removes a packet from the ring
and places it in \code{dst}. Fig.\ref{lst:vigor} shows the resulting NF code.
\code{VIGOR_CHECK} adds additional property that we would like to verify ---
that we do not forward packets with port 9.

@@

{\bf Step 3: Provide invariants.} Besides usual parameters, like capacity or
element size, most Vigor components also require simple logical invariants of
their elements to enable the reasoning about the application code. For our
example, the essential property of the ring is, that it contains only packets
with a port different from 9. A developer communicates that as shown in
Fig.~\ref{lst:customizations}.

@@

Currently, we have to duplicate the packet condition in two languages (formal
logic, and C) due to the limitations of our prototype.

{\bf Step 4: Run Vigor checker.} Vigor will find the appropriate symbolic model,
check that the application does not crash in any execution scenario. It will
then make sure that the chosen model is faithful to the formally proven
implementation of the component.

{\bf Step 5: Link the NF code with the verified library} of components. The
resulting executable is guaranteed to be free of crashes, and hold the
properties formulated in \code{VIGOR_CHECK}, subject to the assumptions listed
in~\S\ref{sec:limitations}.

\subsection{Under the hood}

One could use an SEE like KLEE~\cite{cadar2008klee} to symbolically execute the
original application code (Fig.~\ref{lst:orig}) exhaustively --- once all paths
are found to be crash-free, the crash-freedom proof is complete. To perform SE,
it is necessary to provide symbolic models for the three network calls.
Something simple suffices, such as \code{receive_packet} generating a new
symbolic packet in \code{dst}; \code{can_send_packet} forking program state into
two states for the interface being busy and idle correspondingly; and
\code{send_packet} doing nothing.

Alas, naive symbolic execution would never terminate. The reason is ``path
explosion'': there is a large (often infinite) number of paths to explore, so
the SEE cannot finish. Path explosion results from cyclomatic
complexity~\cite{mccabe:cyclomatic}: every branch condition doubles the number
of paths, and every loop multiplies it by a factor; in our case, the infinite
loop gives rise to an infinite number of paths.

Vigor transforms infinite loops using loop
invariants~\cite[\S~2.1]{cormen2009introduction}. We added to KLEE support for
automatic havoc-ing~\cite{barnett2005boogie} (replace all variables changed in
the loop with unconstrained symbols). The step 1 enables Vigor to reduce each
problematic loop to a single generalized iteration that allows reasoning in one
pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. Another classic SE challenge ---
``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset} spoils
the party here. When
indexing an array with a symbolic variable (or when dereferencing a symbolic
pointer), the constraints on program state become extremely complex, because the
symbolic pointer or index could be referencing many different possible
locations. The complexity of the constraints chokes the SMT solver, and the SEE
grinds to a halt. In general, SE doesn't work for code with symbolic pointers or
offsets. Alas, stateful C code almost always has some of these.

In Vigor, we circumvent the problem of symbolic indexing by encapsulating program
state into well-defined data structures and putting them into a ``library of
components'' (step 2). This works well for NF development, because the NF code
uses a relatively small set of standard structures, such as ring buffers, hash
tables, least-prefix-match tables, etc. We provide to the SEE symbolic models of
these abstracted components that ``hide'' symbolic indexes and pointers from the
SEE.

After the step 2 exhaustive SE takes less than 100 millisec regardless of the
value of \code{CAP}. The fact that the code does not
crash on any path amounts to a proof of crash freedom. Or does it? The validity
of the proof depends on our library implementation of \code{ring} not
crashing; and (2) the symbolic model of \code{ring} accurately capturing how the
\code{ring} implementation works.

The Vigor library comes with a formal specification and a formal proof that the
implementation satisfies that specification. In other words, there is a formal
spec for \code{ring} and a proof that the \code{ring} implementation satisfies
the spec. This is how we enable SE to verify the high level code, while using
separation logic to verify the library of components --- two different
verifications techniques for two different parts of the complete application.
Doing the library verification requires deep expertise in program verification,
but our effort is amortized across all NFs that use the library.

Vigor makes it possible for NF developers to verify their code without having to
understand verification. Note that stating invariants(step 3) for dataplane
software turns out to require no verification expertise.

Here is an example of the proven specification for the function
\code{ring_pop_front}:

@@

\begin{lstlisting}
ringp(r, user_property, tail(lst), cap) &*&
packetp(p, head(lst)) &*&
true == user_property(head(lst));
\end{lstlisting}
means:
\begin{itemize}
\item \code{r} points to a healthy ring of capacity \code{cap}, that contains
  \code{tail(lst)} packets, that satisfy the \code{user_property}.
\item \code{p} points to a packet ``\code{head(lst)}''.
\item ``\code{head(lst)}'' satisfies the \code{user_property}.
\end{itemize}

Note that \code{user_property(head(lst));} in the post condition ensures
\code{p->port != 9} property (Fig.~\ref{lst:customizations}).

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  ASSUME(packet_constraints(p));
}
\end{lstlisting}
  \caption{A symbolic model for \code{ring_pop_front}.}
  \label{lst:symbolic-model}
\end{figure}

Some symbolic models work well for some code, while others work well for other
code, so the Vigor library comes with {\em multiple} simple models for each
component, and Vigor automatically chooses combinations of models. Once it finds
a combination of models with which the desired property (crash freedom, no
memory leaks, absence of certain security vulnerabilities) can be proven, Vigor
validates that the behaviors exposed by the model during verification are
consistent with the component's formal specification. Details on how it does
this appear in \S\ref{sec:design}.

A model of \code{ring_pop_front} used for this example is shown on
Fig.~\ref{lst:symbolic-model}. This model is simple enough to enable
exhaustive SE, but preserves the vital propertiy --- \code{packet_constraints}
(Fig.\ref{lst:customizations}).

@@

You can see one of the generated call sequences on Fig.~\ref{lst:call-sequence}.
It starts with \code{loop_invariant_produce} that marks the beginning of a loop
iteration. It continues then with calls to \code{ring_full} \code{ring_empty},
noting (with \code{assume}) that both return \code{false}. The seqeunce ends
with a call to \code{ring_pop_front}. The last line detects underapproximation
of the model output.

\begin{figure}[h]
\begin{lstlisting}
struct ring* arg1;
struct packet arg2;

@@

/*@ assert(arg2.port != 9); @*/
\end{lstlisting}
  \caption{A call sequence, produced by SEE.}
  \label{lst:call-sequence}
\end{figure}

Formal verification engine checks this sequence using the contracts like in the
Fig.~\ref{lst:contract}, and ensures the abscence of misuse e.g. call for
\code{ring_pop} on a possibly empty ring. It then checks that the model output
covers all the possible outputs specified in the contract for the tip function.
This happens in the last line on the listing. The assert contains the
path-constraint generated by the symbolic model. In this case the contract
(Fig.\ref{lst:contract}) guarantees that the popped packet satisfies
\code{user_property} that entails \code{arg2.port != 9}, hence the trace is
ligitimate. (see~\S\ref{sec:model-validation-criteria}).

Consider potential model mismatches. Suppose we had an overapproximate model ---
Fig.~\ref{lst:overapproximate}. Then the \code{VIGOR_CHECK}
(Fig.\ref{lst:vigor}) in the application code would fail.
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  // No constraints on the packet port here.
}
\end{lstlisting}
  \caption{An overapproximate ring model fragment.}
  \label{lst:overapproximate}
\end{figure}

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  p->port = 0;
}
\end{lstlisting}
  \caption{An underapproximate ring model fragment.}
  \label{lst:underapproximate}
\end{figure}

Suppose now we had an underapproximate model: Fig.~\ref{lst:underapproximate}.
Then the last line in the call sequence check (Fig.\ref{lst:call-sequence}) would look like:
\begin{lstlisting}
/*@ assert(arg2.port == 0); @*/
\end{lstlisting}
The assertion fails, because the contract (Fig.~\ref{lst:contract}) specifies
a wider range for \code{arg2.port} than just \code{0}.
